A Coq Formalization of Stable Sorts, Radix Sort, and the Burrows-Wheeler Transform

### 2019-10-16 draft_paper/draft_paper.tex
stash

\begin{abstract}
Stable sorts are an important and fundamental type of sorting
algorithm. Some algorithms, such as radix sorts, rely upon the
stability of a sorting subroutine for their correctness. In other
cases, stable sorts are useful for reasoning about algorithms, often
because any two stable sorts are interchangeable. One such algorithm
is the Burrows--Wheeler transform (BWT), which has applications in
compression, full-text indexing and pattern searching.

There has not been much work on formally verifying stable sorts, or
rather, the stability of stable sorts. While Bird
et~al.\cite{birdmu,pearls} provide an elegant derivation and proof for
the BWT, formalizing it in Coq reveals that it relies on a few lemmas
that are not trivial to prove, including the correctness of
least-significant digit (LSD) radix sort. In this paper, we develop
more fully the theory of stable sorts, and use this theory to prove
the correctness of LSD radix sort and thereby the BWT.
\end{abstract}

@@

Stability can be directly useful in certain contexts, such as when
sorting a table by a single column. However, stability is also crucial
to the correctness of certain general-purpose algorithms. One such
algorithm, or set of algorithms, is radix sort. Radix sorts are used
to sort data lexicographically, and can be more efficient than
comparison-based sorting methods, both in theory and in practice.
Radix sorts are especially well-suited to sorting strings
\cite{McIlroy93}. Many radix sorts rely on a some stable sorting
subroutine.

@@

In addition, to our knowledge no one has produced a formally verified
implementation of the BWT, by which we mean an implementation with a
proof that the inverse-transform inverts the forwards-transform. Bird
et~al.\cite{alg-of-prog,birdmu} provide an excellent derivation of the
BWT and its inverse that we use as the basis of our implementation and
proof. However, it leaves a few crucial lemmas unproven, including the
correctness of LSD radix sort.

In this paper, we develop more fully the theory of stable sorts: we
give multiple definitions for permutations and stable permutations and
prove them equivalent. This allows us to choose whichever definition
makes the proof easiest when we are reasoning about stability. We also
proved that any two sorts are interchangeable.
\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
  \label{thm:stablesort_unique}
  For all functions $f, g: \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{StableSort}~f \implies \var{StableSort}~g \implies
    (\forall l,f~l = g~l)
  \end{equation*}
\end{restatable*}

Using \cref{thm:stablesort_unique}, we prove the correctness of an LSD
radix sort that relies on insertion sort to sort each column. This
choice was crucial to the proof: parametrizing radix sort over any
stable sort made it difficult to reason about, but reasoning about
insertion sort using induction is very elegant.
\cref{thm:stablesort_unique} allows us to recover the generality of
parametrizing over all stable sorts, because insertion sort is
equivalent to any other stable sort.

Finally, we use the correctness of radix sort to reason about the BWT.
Specifically, we implement the functions $\var{bwl} : \var{list}~A \to
\var{list}~A$ and $\var{bwn} : \var{list}~A \to \mathbb{N}$ that produce
the string and index part of the BWT respectively, and the function
$\var{unbwt} : \mathbb{N} \to \var{list}~A \to \var{list}~A$ that reverses
the transform. We then prove the following theorem:
\begin{restatable*}[unbwt\_correct]{theorem}{unbwtcorrect}
  \label{thm:unbwt_correct}
  For all lists $l$,
  \begin{equation*}
    \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
  \end{equation*}
\end{restatable*}

@@

\section{Main ideas}
\label{sec:main_ideas}

\subsection{Stable sorting}
\label{subsec:stable_sorting}

Whenver we sort a list, we do so with respect to some ordering on its
elements. For example, we could sort a list of students by first name,
last name, birthday, etc. More precisely, for sorting we need a total,
decideable preorder: a \textit{preorder} is a relation $\le$ that is
reflexive and transitive, it is \textit{total} if we can compare any
two elements, and it is \textit{decidable} if we have a procedure that
takes any two elements $x$ and $y$ and tells us if $x \le y$ or $x \ge y$.

Given a total decideable preorder, we can specify what contract a
function~$f$ must fulfill to be considered a sorting function.
Intuitively, it must produce a list that is sorted, and that contains
all of the same elements as the original list.
\begin{definition}[Sort]
  A function $f : \var{list}~A \to \var{list}~A$ is a \var{Sort} when
  for all inputs $l$,
  \begin{equation*}
    \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
  \end{equation*}
\end{definition}

Now we have to define \var{Sorted} and \var{Permutation}
appropriately. For both properties, and for other properties we will
discuss later, there are multiple reasonable definitions. We will
state all the definitions we can think of, and prove equivalence
between them. This increases our confidence that we have defined
things correctly, and allows us to choose whichever definition we find
most convenient for a given proof.

We can define \var{Sorted} in two main ways: %TODO cite coq book and knuth
\begin{enumerate}
\item A list $l$ is sorted when for all
  indices $i$ and $j$,
\begin{equation*}
  i \le j \implies l_i \le l_j
\end{equation*}
\item We can inductively define when a list $l$ is sorted, by saying
  that the empty list is sorted, and that given a sorted lists $t$, we
  may conclude that $h :: t$ is also sorted if $h$ is less~than or
  equal~to every element in $t$.
\end{enumerate}

Similarly, we can define \var{Permutation} in three main ways:
\begin{enumerate}
\item Two lists $l$ and $l'$ are permutations of one another if every
  $x$ occurs an equal number of times in $l$ and $l'$.
\item Two lists $l$ and $l'$ with lengths $n$ are permutations of one
  another when there exists a permutation function~$p$ that maps the
  indices of $l$ to indices of $l'$, so that for every index~$i$
  \begin{equation*}
      l_i = l'_{p(i)}
  \end{equation*}
  $p$ must be a bijection so that every index of $l$ is mapped to by
  exactly one index of $l'$.
\item We can define inductively when two lists are permutations of one another:
  \begin{itemize}
  \item As a base case, two empty lists are permutations of one
    another.
  \item If two lists $l$ and $l'$ are permutations of one another,
    then we can conclude that $x :: l$ and $x :: l'$ are permutations
    of one another.
  \item As a second base case, we can conclude that for any $x$, $y$,
    and $l$, that $x :: y :: l$ is a permutation of $y :: x :: l$.
  \item If $l_1$ is a permutation of $l_2$ and $l_2$ is a permutation
    of $l_3$, then $l_1$ is a permutation of $l_3$.
  \end{itemize}
\end{enumerate}

Now that we've defined what a sort is, we can start to consider stable
sorts. Stable sorts must preserve the relative orders of
``equivalent'' elements. This equivalence comes from the ordering we
are using: any total decidable preorder also give us a decideable
equivalence relation $\equiv$, where $x \equiv y$ when $x \le y$ and $x \ge y$.
\begin{definition}[StableSort]
  A function $f : \var{list}~A \to \var{list}~A$ is a \var{StableSort} when
  for all inputs $l$,
  \begin{equation*}
    \var{Sorted}~(f~l) \land \var{StablePerm}~(f~l)~l
  \end{equation*}
\end{definition}

Stable sorts still must produce sorted outputs, but the output cannot
be just any permutation of the input. So we need a stricter notion of
a permutation that respects equivalence, which we will call a stable
permutation.

To define stable permutations, we can tweak the each
definitions of permutation mentioned above:
\begin{enumerate}
\item Two lists $l$ and $l'$ are stable permutations of one another if
  for every $x$,
  \begin{equation*}
    \var{filter}~(\lambda y. y \equiv x)~l = \var{filter}~(\lambda y. y \equiv x)~l'
  \end{equation*}
  This is a stricter version of the first definition of permutation
  above, because that can be rephrased as $\var{filter}~(\lambda y. y = x)~l
  = \var{filter}~(\lambda y. y = x)~l'$.
\item For two lists $l$ and $l'$ of length $n$ to be stable
  permutations of one another, there must exists a \textit{stable}
  permutation function~$p$ that relates them. That means that $p$
  satisfies an additional stability property: for all indices $i < j$
  where $l_{i} \equiv l_{j}$, we must have $p(i) < p(j)$.
\item We can modify the third inference rule, which allows swapping
  $x$ and $y$, so that you can only apply it when $x \nequiv y$: the new
  rule says that for any $x$, $y$, and $l$ such that $x \nequiv y$ , you can
  conclude that $x :: y :: l$ is a permutation of $y :: x :: l$.
\end{enumerate}

From these definitions, we can see that stable sorts are restricted in
which permutation they produce. In fact, there is only one stable
permutation that is also sorted, so any two stable sorts must produce
the same results for every input, and are thus interchangeable.
\stablesortunique

This can be generalized a little further: even if a sorting algorithm
itself is not a stable sorting algorithm, if the ordering we are using
doesn't allow us to distinguish between equivalent elements, then we
will not be able to tell if equivalent elements are being swapped.
More precisely, notice that it isn't always true that $x \equiv y$ implies
$x = y$. For example, if we are sorting strings by their first
character, then $\text{\texttt{apple}} \equiv \text{\texttt{armpit}}$ under
this ordering, but clearly $\text{\texttt{apple}} \neq
\text{\texttt{armpit}}$. When we do have that the equivalence ($\equiv$)
implies Leibniz equality ($=$), we say that we have an \textit{order},
as we have when we are sorting strings using all of their characters.
Under an order, as opposed to a preorder, we can prove that every
permutation is a stable permutation and every sort is a stable sort,
which implies that every sort is equivalent to every other sort.
\begin{theorem}[all\_perm\_stable]
  If we have an equivalence that implies Leibniz equality, then for
  all lists $l$ and $l'$
  \begin{equation*}
      \var{Permutation}~l~l' \implies \var{StablePerm}~l~l'
  \end{equation*}
\end{theorem}
\begin{corollary}[Sort\_Ord\_unique]
  \label{thm:sort_ord_unique}
  When sorting using an order, given any two function $f, g:
  \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{Sort}~f \implies \var{Sort}~g \implies
    (\forall l, f~l = g~l)
  \end{equation*}
\end{corollary}

\subsection{LSD radix sort}
\label{sec:radix_sort}

LSD radix sort is a simple radix sort that is used to
lexicographically sort a list of strings of equal length. LSD radix
sort operates by stably sorting on each column, from the end of the
string to the front (\cref{fig:lsd_radixsort}). The algorithm works
because lexicographic order says that if two strings are equal up to a
column, then the tie is broken according to the rest of the string.
When we stably sort on a column, ties are broken in favor or the
existing order, which was established by sorted on the following
columns.

\begin{figure}[!hb]
  \centering
  \begin{tt}
    \begin{tabular}{c}
      d a \textbf{b} \\
      a d \textbf{d} \\
      c a \textbf{b} \\
      f a \textbf{d} \\
      f e \textbf{e} \\
      b a \textbf{d} \\
      d a \textbf{d} \\
      b e \textbf{e} \\
      f e \textbf{d} \\
      b e \textbf{d} \\
      e b \textbf{b} \\
      a c \textbf{e}
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c}
      d \textbf{a} b \\
      c \textbf{a} b \\
      e \textbf{b} b \\
      a \textbf{d} d \\
      f \textbf{a} d \\
      b \textbf{a} d \\
      d \textbf{a} d \\
      f \textbf{e} d \\
      b \textbf{e} d \\
      f \textbf{e} e \\
      b \textbf{e} e \\
      a \textbf{c} e
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c}
      \textbf{d} a b \\
      \textbf{c} a b \\
      \textbf{f} a d \\
      \textbf{b} a d \\
      \textbf{d} a d \\
      \textbf{e} b b \\
      \textbf{a} c e \\
      \textbf{a} d d \\
      \textbf{f} e d \\
      \textbf{b} e d \\
      \textbf{f} e e \\
      \textbf{b} e e
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c}
      a c e \\
      a d d \\
      b a d \\
      b e d \\
      b e e \\
      c a b \\
      d a b \\
      d a d \\
      e b b \\
      f a d \\
      f e d \\
      f e e
    \end{tabular}
  \end{tt}
  \caption{LSD Radixsort}
  \label{fig:lsd_radixsort}
\end{figure}

\subsection{The Burrows-Wheeler transform}
\label{subsec:bwt}

The Burrows--Wheeler transform was originally introduced as a pass in a
compression algorithm \cite{bw}, and continues to be used as a part of
the popular, open-source bzip2 compression software \cite{tsai_2016}.
Here we describe a standard compression algorithm built around BWT.
\begin{enumerate}
\item Apply the BWT, which tends to group create runs of identical characters.
\item Apply Move-to-Front (MTF) encoding, in which a character is
  represented by the number of different characters encountered since
  the last occurrence of this character. This creates many runs of
  zeroes from the runs in the output of BWT. On English texts, often
  50-60\% of the values produced by the MTF pass are zeroes
  \cite{fenwick2007}.
\item Apply Run-Length Encoding (RLE) to encode only the runs of
  zeroes with their lengths, using a bijective base-2 encoding
  \cite{bw-analysis, tsai_2016}.
\item Compress with some zeroth-order coder like a Huffman coder.
\end{enumerate}

>>>>

\begin{abstract}
Stable sorts are an important and fundamental type of sorting
algorithm. Some algorithms, such as radix sorts, rely upon the
stability of a sorting subroutine for their correctness. In other
cases, stable sorts are useful for reasoning about algorithms, often
because any two stable sorts are interchangeable. One such algorithm
is the Burrows--Wheeler transform (BWT), which has applications in
compression, full-text indexing, and pattern searching.

There has not been much work on formally verifying stable sorts, or
rather, the stability of stable sorts. While Bird et~al.
\cite{birdmu,pearls} provide an elegant derivation and proof for the
BWT, formalizing it in Coq reveals that they rely on a few lemmas that
are not trivial to prove, including the correctness of
least-significant digit (LSD) radix sort. We develop the theory of
stable sorts, and use this theory to prove the correctness of LSD
radix sort and thereby the BWT.
\end{abstract}

@@

Stability can be directly useful in certain contexts, such as when
sorting a table by a single column. However, stability is also crucial
to the correctness of certain general-purpose algorithms. One such
algorithm, or set of algorithms, is radix sort. Radix sorts are used
to sort data lexicographically, and can be more efficient than
comparison-based sorting methods, both in theory and in practice
\cite{McIlroy93,Bentley:1997:FAS:314161.314321,10.1007/978-3-540-89097-3_3}.
Many radix sorts rely on a some stable sorting subroutine.

@@

In addition, to our knowledge no one has produced a formally verified
implementation of the BWT, by which we mean an implementation with a
proof that the inverse-transform inverts the forwards-transform. Bird
et~al.\cite{alg-of-prog,birdmu} provide an excellent derivation of the
BWT and its inverse based on the technique program calculation, in
which programs are calculated from their specifications using
equational reasoning. However, their derivation leaves a few crucial
lemmas unproven, including the correctness of LSD radix sort.

In this paper, we develop more fully the theory of stable sorts: we
give multiple definitions for permutations and stable permutations and
prove them equivalent. This allows us to choose whichever definition
makes the proof easiest when we are reasoning about stability. We also
proved that any two sorts are interchangeable.
\begin{restatable*}[StableSort\_unique]{corollary}{stablesortunique}
  \label{thm:stablesort_unique}
  For all stable sorting functions $f, g: \var{list}~A \to
  \var{list}~A$,
  \begin{equation*}
    (\forall l,f~l = g~l)
  \end{equation*}
\end{restatable*}

Using \cref{thm:stablesort_unique}, we prove the correctness of an LSD
radix sort that relies on insertion sort to sort each column. This
choice was crucial to the proof: parametrizing radix sort over any
stable sort made it difficult to reason about, but reasoning about
insertion sort is very elegant. \cref{thm:stablesort_unique} allows us
to recover the generality of parametrizing over all stable sorts,
because it proves that insertion sort is equivalent to any other
stable sort.

Finally, we use the correctness of radix sort to reason about the BWT.
Specifically, we implement the functions $\var{bwl} : \var{list}~A \to
\var{list}~A$ and $\var{bwn} : \var{list}~A \to \nat$ that produce
the string and index part of the BWT respectively, and the function
$\var{unbwt} : \nat \to \var{list}~A \to \var{list}~A$ that reverses
the transform. We then prove the following theorem:
\begin{restatable*}[unbwt\_correct]{theorem}{unbwtcorrect}
  \label{thm:unbwt_correct}
  For all lists $l$,
  \begin{equation*}
    \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
  \end{equation*}
\end{restatable*}

@@

\section{Sorting}
\label{sec:sorting}

Whenever we sort a list, we do so with respect to some ordering on its
elements. For example, we could sort a list of students by first name,
last name, birthday, etc. More precisely, for sorting we need a total,
decideable preorder: a \textit{preorder} is a relation $\le$ that is
reflexive and transitive, it is \textit{total} if we can compare any
two elements, and it is \textit{decidable} if we have a procedure that
takes any two elements $x$ and $y$ and tells us if $x \le y$ or $x \ge y$.

Given a total decideable preorder, we can define when a list is in
ascending order with respect to that preorder. In this case, as will
be the case with other properties we will discuss, there are multiple
reasonable definitions. Our approach will be to state all the
definitions we can think of and to prove equivalence between them.
This increases our confidence that we have defined things correctly,
and allows us to choose whichever definition we find most convenient
for a given proof.

We can define \var{Sorted} in two main ways: an index-based definition
(\var{SortedIx}), and an inductive definition (\var{Sorted}).
\begin{definition}{SortedIx}
  A list $l$ is sorted if for all indices $i$ and $j$,
  \begin{equation*}
    i \le j \implies l_i \le l_j
  \end{equation*}
\end{definition}

\begin{definition}{Sorted}
  \begin{align*}
    \infer[\var{Sorted\_nil}]{\var{Sorted}~[]}{}
    &&
    \infer[\var{Sorted\_cons}]{\var{Sorted}~(a :: l)}{%
      (\forall~x \in l,~a < x) & \var{Sorted}~l
    }
  \end{align*}
\end{definition}

\begin{restatable}[SortedIx\_iff]{theorem}{sortedixiff}
  For all lists $l$,
  \begin{equation*}
    \var{Sorted}~l \iff \var{SortedIx}~l
  \end{equation*}
\end{restatable}
\begin{proof}
  See \cref{appendix:proofs}.
\end{proof}

However, a sorting function cannot produce just any sorted list as its
output; the sorted list must also contain all the same elements as the
original list. To capture this, we could use the \var{Permutation}
relation defined in the Coq standard library. This relation holds on
two lists when they contain all the same elements, and is defined
inductively:
\begin{definition}[Permutation]
  \label{def:permutation}
  \begin{gather*}
    \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
    \\
    \allowbreak
    \\
    \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
      \var{Permutation}~l~l'
    }
    \\
    \allowbreak
    \\
    \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
    \\
    \allowbreak
    \\
    \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
      \var{Permutation}~l~l' & \var{Permutation}~l'~l''
    }
  \end{gather*}
\end{definition}

Before we go further, we should clarify what we mean by the word
\textit{permutation}, as it has two related meanings that are often
used interchangeably. We sometimes say that two lists are permutations
of one other if they contain all of the same elements, perhaps in a
different order, like the lists $l =
[\var{zebra},\ \allowbreak\var{appliance},\ \allowbreak\var{kite}]$
and $l' =
[\var{appliance},\ \allowbreak\var{kite},\ \allowbreak\var{zebra}]$
However, in mathematics we sometimes say that a permutation is
actually the bijection on the indices that defines the correspondence
between the two lists. In this case, we could map indices in $l$ to
indices in $l'$ by mapping $0 \mapsto 2$, mapping $1 \mapsto 0$, and mapping $2 \mapsto
0$. Formally, we say that $p$ is the permutation that relates $l$ to
$l'$ if for all indices $i$ and $j$,
\begin{equation*}
  l_i = l'_{p(i)}
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
one other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$.

This suggests an alternative definition for \var{Permutation}:
\begin{definition}{PermutationEx}
  \label{def:permutation_ex}
  Two lists $l$ and $l'$ are permutations of one another when, letting
  $n = \var{length}~l$, there exists a permutation function $p$ on $n$
  elements such that $p(l) = l'$.
\end{definition}
Here we are overloading the notation $p(l)$ to denote the result of
shuffling a list~$l$ according to a permutation function~$p$. To
define this precisely, we have to first define how we will represent
permutation functions in Coq.

A permutation function on $n$ elements maps indices to indices, and so
has type $\nat_{<n} \to \nat_{<n}$. In Coq, we can represent
a finite function as a list of its outputs: $f : \nat_{<n} \to A$
is represented by the list $[f(0),~f(1),\allowbreak\ldots,~f(n-1)]$.
Applied to permutations, this means that the permutation~$p$ mentioned
above would be represented by $[2,~1,~0]$. This is consistent with
``one-line'' notation for permutations, in which a permutation on $n$
elements is represented by its result when applied to the
list~$[0,~2,\ldots,n-1]$. Now we can make precise the notation
$p(l)$\footnote{In Coq, this corresponds the the \var{apply} function,
  whose definition is slightly more complex due to the extra default
  parameter for \var{nth}.}
\begin{equation*}
  p(l) \coloneq \var{map}~(\lambda i. l_i)~p
\end{equation*}

However, not any list of natural numbers represents a valid
permutation function, because a permutation must be a bijection. We
can define the property $\var{PermFun}~n~p$, which holds when $p :
\var{list}~\nat$ represents a permutation function on lists of
length $n$:
\begin{definition}[PermFun]
  A list~$p$ of natural numbers is a permutation function on $n$
  elements if for all natural numbers $i$, $i \in p \iff i < n$, and if
  $p$ contains no duplicate elements\footnote{For this we can use the
    \var{NoDup} predicate from Coq's standard library.}.
\end{definition}

We should also mention a third definition for permutations that can be
used on types with decideable equality:
\begin{definition}[PermutationCount]
  \label{def:permutation_count}
  Two lists are permutations of one another when for any $x$,
  \begin{equation*}
    \var{count\_occ}~l~x = \var{count\_occ}~l'~x
  \end{equation*}
  where $\var{count\_occ}~l~x$ counts the number of elements in $l$
  that are equal to $x$.
\end{definition}

With this we can define the specification for sorting functions:
\begin{definition}[Sort]
  A function $f : \var{list}~A \to \var{list}~A$ is a sorting function
  when for all inputs $l$,
  \begin{equation*}
    \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
  \end{equation*}
\end{definition}

\section{Stable sorting}
\label{sec:stable_sorting}

Now that we've defined sorts, we can start to consider stable sorts.
Stable sorts must preserve the relative orders of ``equivalent''
elements. This equivalence comes from the ordering we are using: any
total decidable preorder also give us a decideable equivalence
relation $\equiv$, where $x \equiv y$ when $x \le y$ and $x \ge y$.

Stable sorts still must produce sorted outputs, but the output cannot
be just any permutation of the input. We need a stricter notion of
a permutation that respects equivalence, which we will call a stable
permutation.
\begin{definition}[StableSort]
  A function $f : \var{list}~A \to \var{list}~A$ is a stable sorting
  function when for all inputs $l$,
  \begin{equation*}
    \var{Sorted}~(f~l) \land \var{StablePerm}~(f~l)~l
  \end{equation*}
\end{definition}

To define stable permutations, we can tweak each of the definitions of
permutation mentioned above. For example, we can modify our
function-based definition.

\begin{definition}[StablePermEx]
  Two lists $l$ and $l'$ are stable permutations of one another when
  there exists a \textit{stable} permutation function~$p$ on $l$ such
  that $p(l) = l'$.
\end{definition}

\begin{definition}[StablePermFun]
  A permutation function $p$ is a \textit{stable} permutation function
  on some list $l$ when for all indices $i < j$ where $l_i \equiv l_j$, we
  also have $p(i) < p(j)$.
\end{definition}

We can also modify the inductive definition of permutations to respect
equivalence. The problematic rule is \var{perm\_swap} because we could
porentially swap two equivalent elements. We will restrict it, so that
you can only swap $x$ and $y$ when $x \nequiv y$.

\begin{definition}[StablePermInd]
  \begin{gather*}
    \infer[\var{StablePermInd\_nil}]{\var{StablePermInd}~[]~[]}{}
    \\ \\
    \infer[\var{StablePermInd\_skip}]{\var{StablePermInd}~(x :: l)~(x :: l')}{%
      \var{StablePermInd}~l~l'
    }
    \\ \\
    \infer[\var{StablePermInd\_swap}]
          {\var{StablePermInd}~(x :: y :: l)~(y :: x :: l)}
          {x \not\equiv y}
    \\ \\
    \infer[\var{StablePermInd\_trans}]{\var{StablePermInd}~l~l''}{
      \var{StablePermInd}~l~l' & \var{StablePermInd}~l'~l''
    }
  \end{gather*}
\end{definition}

Finally, we can say that if two lists are in a stable permutation, and
we examine only the the elements equivalent to $x$ for some $x$, then
those elements should appear in the same order in both lists.
\begin{definition}[StablePerm]
\item Two lists $l$ and $l'$ are stable permutations of one another if
  for every $x$,
  \begin{equation*}
    \var{filter}~(\lambda y. y \equiv x)~l = \var{filter}~(\lambda y. y \equiv x)~l'
  \end{equation*}
\end{definition}
In a sense, this is a stricter version of
\cref{def:permutation_count}, because that can be rephrased as
$\var{filter}~(\lambda y. y = x)~l = \var{filter}~(\lambda y. y = x)~l'$.

From these definitions, we can see that stable sorts are restricted in
which permutation they produce. In fact, there is only one stable
permutation that is also sorted.
\begin{theorem}[StablePerm\_Sorted\_eq]
  For all lists $l$,
  \begin{equation*}
    \var{Sorted}~l \implies \var{Sorted}~l' \implies
    \var{StablePerm}~l~l' \implies l = l'.
  \end{equation*}
\end{theorem}
\begin{proof}
  We reason by induction on $l$. In the inductive case, we have to
  show that
  \begin{equation*}
    h :: t = h' :: t'
  \end{equation*}
  given $\var{Sorted}~(h :: t)$, $\var{Sorted}~(h' :: t')$, and
  $\var{StablePerm}~(h :: t)~(h' :: t')$, as well as $t = t'$ from the
  induction hypothesis. So it suffices to show $h = h'$.

This means that any two stable sorts must produce the same results for
every input, and are thus interchangeable.
\stablesortunique*

This can be generalized a little further: even if a sorting algorithm
itself is not a stable sorting algorithm, if the ordering we are using
doesn't allow us to distinguish between equivalent elements, then we
will not be able to tell if equivalent elements are being swapped.
More precisely, notice that it isn't always true that $x \equiv y$ implies
$x = y$. For example, if we are sorting strings by their first
character, then $\text{\texttt{apple}} \equiv \text{\texttt{armpit}}$ under
this ordering, but clearly $\text{\texttt{apple}} \neq
\text{\texttt{armpit}}$. When we do have that the equivalence ($\equiv$)
implies Leibniz equality ($=$), we say that we have an \textit{order},
as we have when we are sorting strings using all of their characters.
Under an order, as opposed to a preorder, we can prove that every
permutation is a stable permutation and every sort is a stable sort,
which implies that every sort is equivalent to every other sort.

\begin{corollary}[Sort\_StableSort\_Ord]
  \label{thm:stablesort_ord}
  When sorting using an order,
  \begin{equation*}
    \var{Sort}~f \iff \var{StableSort}~f
  \end{equation*}
  for any $f$.
\end{corollary}

\begin{corollary}[Sort\_Ord\_unique]
  \label{thm:sort_ord_unique}
  When sorting using an order, for any two sorting function $f, g:
  \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    (\forall l, f~l = g~l)
  \end{equation*}
\end{corollary}

\section{LSD radix sort}
\label{sec:radix_sort}

LSD radix sort is a simple radix sort that is used to
lexicographically sort strings of equal length. We can define what it
means to sort lexicographically:

\begin{definition}[lex\_le]
  Given a preorder~$\le$ on a type $A$, we can define another preorder,
  $\lexle$ on $\var{list}~A$.
  \begin{gather*}
    \infer[\var{lex\_le\_nil}]{[] \lexle ys}{}
    \\
    \\
    \infer[\var{lex\_le\_cons\_lt}]{(x :: xs) \lexle (y :: ys)}{%
      x < y
    }
    \\
    \\
    \infer[\var{lex\_le\_cons\_eq}]{(x :: xs) \lexle (y :: ys)}{%
      x \equiv y & xs \lexle ys
    }
  \end{gather*}
\end{definition}

LSD radix sort operates by conceptually placing all the strings in a
matrix, and by stably sorting on each column, from the end of the
string to the front (\cref{fig:lsd_radixsort}). This works because if
two strings match on a column, the stable sort ``defers'' to the
preexisting ordering, and that preexisting ordering was determined by
sorting by the tail of the strings. This mirrors the way a
lexicographic ordering ``defers'' to the rest of a string when two
strings match on earlier characters.

\begin{figure}
  \centering
  \begin{tt}
    \setlength{\tabcolsep}{0pt}
    \begin{tabular}{c}
    fit \\
    rib \\
    fly \\
    vat \\
    age \\
    mix \\
    row \\
    fog
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    ri&b& \\
    ag&e& \\
    fo&g& \\
    fi&t& \\
    va&t& \\
    mi&x& \\
    fl&y& \\
    ro&w&
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    v&a&t \\
    a&g&e \\
    r&i&b \\
    f&i&t \\
    m&i&x \\
    f&l&y \\
    f&o&g \\
    r&o&w
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &a&ge \\
    &f&it \\
    &f&ly \\
    &f&og \\
    &m&ix \\
    &r&ib \\
    &r&ow \\
    &v&at
    \end{tabular}
  \end{tt}
  \caption{LSD radix in-place}
  \label{fig:lsd_radixsort}
\end{figure}

In a functional setting, we can define an LSD radixsort elegantly by
repeatedly moving the last column to the front of the matrix and
sorting on the first column (\cref{fig:lsd_radixsort_func}). We will
take as a parameter the number of columns $n$.
\begin{equation*}
  \var{radixsort}~m~n = (\var{hdsort} \comp \var{map}~\var{rrot})^n~m
\end{equation*}

\begin{figure}
  \centering
  \begin{tt}
    \setlength{\tabcolsep}{0pt}
    \begin{tabular}{c}
    fit \\
    rib \\
    fly \\
    vat \\
    age \\
    mix \\
    row \\
    fog
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c}
    tfi \\
    bri \\
    yfl \\
    tva \\
    eag \\
    xmi \\
    wro \\
    gfo
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &b&ri \\
    &e&ag \\
    &g&fo \\
    &t&fi \\
    &t&va \\
    &x&mi \\
    &y&fl \\
    &w&ro
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    ibr \\
    gea \\
    ogf \\
    itf \\
    atv \\
    ixm \\
    lyf \\
    owr
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &a&tv \\
    &g&ea \\
    &i&br \\
    &i&tf \\
    &i&xm \\
    &l&yf \\
    &o&gf \\
    &o&wr
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    vat \\
    age \\
    rib \\
    fit \\
    mix \\
    fly \\
    fog \\
    row
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &a&ge \\
    &f&it \\
    &f&ly \\
    &f&og \\
    &m&ix \\
    &r&ib \\
    &r&ow \\
    &v&at
    \end{tabular}
  \end{tt}
  \caption{LSD radix using rotations}
  \label{fig:lsd_radixsort_func}
\end{figure}

Here \var{hdsort} is an insertion sort that only examines the first
character of each string. We can implement this by defining orderings
that compare elements using some key function:

\begin{definition}[keyOrd]
  Given a preorder on a type $K$ and a key function $\var{key} : A \to
  K$, we can define a preorder on $A$ that compares all elements using
  the key function. That is, we can define the relation
  $\le_{\var{key}}$ on $A$ such that for all $x, y : A$
  \begin{equation*}
    x \le_{\var{key}} y \coloneqq (\var{key}~x) \le (\var{key}~y)
  \end{equation*}
\end{definition}

Now we can define prefix lexicographic orderings on lists, where the
lexicographic ordering is only allowed to compare up to the first $n$
elements of two lists. We define the relation $\le_n$ on $\var{list}~A$
such that for all $xs, ys : \var{list}~A$,
\begin{equation*}
  xs \le_n ys \coloneqq (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
\end{equation*}
We also define $\var{PrefixSorted}~n$ as \var{Sorted} specialized to
the $\le_n$ ordering. \var{hdsort} is insertion sort specialized to the
$\le_1$ ordering.

We would like to prove that radix sort is a stable sort.

\begin{corollary}[radixsort\_correct]
  \label{thm:radixsort_correct}%
  For all matrices $m$ with $n$ columns,
  \begin{equation*}
    \var{Sorted}~(\var{radixsort}~m~n) \land \var{StablePerm}~(\var{radixsort}~m~n)~m
  \end{equation*}
\end{corollary}

For both properties, our approach will be to prove some invariant that
holds after $j$ iterations. For \var{StablePerm}, we can say that
after $j$ iterations of rotating and sorting, the matrix should be a
stable permutation of the original matrix had we rotated $j$ times.

\begin{restatable}[radixsort\_stable\_inv]{theorem}{radixsortstableinv}
  \label{thm:radixsort_stable_inv}
  For all matrices $m$ and $j : \nat$,
  \begin{equation*}
    \var{StablePerm}~(\var{radixsort}~m~j)~((\var{map}~\var{rrot})^j~m)
  \end{equation*}
\end{restatable}

For \var{Sorted}, the invariant will be that after $j$ iterations, the
matrix is sorted on the first $j$ columns.

\begin{restatable}[radixsort\_sorted\_inv]{theorem}{radixsortsortedinv}
  \label{thm:radixsort_sorted_inv}
  For all matrices $m$ with rows of length $n$, and $j \le n$,
  \begin{equation*}
    \var{PrefixSorted}~j~(\var{radixsort}~m~j)
  \end{equation*}
\end{restatable}

To prove both of these invariants, we will have to be able to reason
about the relationship between the various orderings we are using: we
use $\le_1$ in \var{hdsort}, $\le_j$ in $\var{PrefixSorted}~j$, and
$\lexle \approx \le_n$ everywhere else.

Let's generalize, and consider two orders $\le_j$ and $\le_k$ such that $j
\le k$. If in the first $j$ elements we can conclude that $x <_j y$,
then examining more elements can't change that.
